Cluster Difference Imaging Photometric Survey. III. An Open Cluster Spread Across 500\,pc

### 2021-03-09 cluster_paper/ms.tex
improved. manuscript.
* sanity checking my coordinate systems
* adding sick vectors
* add v_tangential check. yeah, it does matter.

\title{
  Cluster Difference Imaging Photometric Survey. III.
  %The Coeval Core \& Halo of NGC\,2516
  %The Halo of NGC\,2516 Extends To At Least 300\,pc
  An Open Cluster With a 300\,pc Halo
  %NGC\,2516  300\,pc
}

@@

% NB. the "classical tidal radius" is from e.g., Piskunov+08 or
% Kharchenko+2013. The latter givs a radius r2 of 0.71 degrees. At 380
% pc, corresponds to 4.7pc. The King tidal radius given by K13 is
% 7.70pc.
\begin{abstract}
  Recent analyses of the Gaia data have reported the existence of
  diffuse stellar populations surrounding nearby open clusters. 
  The stars in these diffuse halos could be the tidal escapees of the
  cluster cores, or they could be remnants of an initially dispersed
  star formation complex.
  They could also be field stars---the false positives of overly
  aggressive clustering algorithms.
  In this study, we focus on a halo with a length of 300\,pc reported
  around the $\approx$150\,Myr open cluster NGC\,2516.
  The classical tidal radius of the cluster is $\lesssim$10\,pc.
  Combining photometry from Gaia, rotation periods from TESS, and
  lithium measurements from Gaia-ESO and GALAH, we find that the halo
  of NGC\,2516 is real and coeval with the core.
  Two in three kinematically selected halo members 25-250\,pc from the
  cluster center show rotation periods consistent with a
  gyrochronological age of 120\,Myr.
  A comparison sample of field stars shows no such trend.
  The rotation period and lithium abundances of the cluster stars are
  also anti-correlated at fixed stellar mass, as has been observed in
  similar stellar populations.
  This work expands the set of confirmed NGC\,2516 members by a factor
  of $\approx$2, and quantifies the degree to which Gaia-based
  analyses can identify dispersed stellar populations against a
  background of field star contaminants.
  We highlight the implications for spectroscopic survey targeting,
  open cluster dispersal, and planet searches around young stars.
\end{abstract}

\keywords{
	stellar ages (1581),
  stellar associations (1582),
  open clusters (1160),
  %stellar dynamics (1596),
  stellar rotation (1629)
}

@@

Star clusters form from the densest filaments of 
hierarchically structured molecular clouds
\citep{shu_star_1987}.  Over the first 10\,Myr, feedback effects
including protostellar outflows, photoionization, radiation pressure,
and supernova shocks disperse the gas out of the cloud
\citep{krumholz_star_2019}.  Since only $\sim 1\%$ of the cloud
mass is converted into stars, gas dispersal enables the
majority ($\sim$90\%) of stars in the cluster to escape the cluster's
gravitational well \citep{lada_embedded_2003}.

@@

%TODO: double check this owrding.
A brief note on terminology.  Low-density stellar associations
connected to a dense population (the ``core'') have been described as
being in ``halos'', ``strings'', ``coronae'', ``snakes'',
``outskirts'', and ``tidal tails'' \citep[{\it
e.g.},][]{davenport_death_2010,kounkel_untangling_2019,roser_hyades_2019,tian_discovery_2020,meingast_2021}.
The latter term implies a particular model for the formation of the
association.  We opt for ``halo'' because it is concise and model
agnostic.  It suffers however from the suggestion of spherical
symmetry.  The halo of NGC\,2516 might be best-described as ``tails'',
but the primary goal of this study is
not to verify whether these tails are tidal.

@@

@@

@@

@@

\begin{figure*}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.95\textwidth]{f4.pdf}
	\end{center}
	\vspace{-0.7cm}
	\caption{ {\bf Gaia-based components of NGC\,2516 in position and
    velocity space, cross-matched against the rotators.}
    The plotted stars are those with $0.5<(Bp-Rp)_0<1.2$ that meet the
    crowding restrictions described in Section~\ref{subsec:tess}: they
    should have been sufficiently bright and non-crowded to enable
    rotation period detection.
		\label{fig:gaia6d_x_rotn}
	}
\end{figure*}

@@

\begin{figure}[t]
	\begin{center}
			\includegraphics[width=0.49\textwidth]{f8b.pdf}
	\end{center}
	\vspace{-0.7cm}
  \caption{ {\bf Histogram of Figure~\ref{fig:physical_x_rotn}, binned
  versus 3d separation from the core and 2d velocity difference.} Bins have width
  25\,pc in position, and 2.0\,km\,s$^{-1}$ in velocity.
  {\bf Uncertainties in the fraction are taken as the quadrature sum of
  the rel unc of the numerator and denominator (FIXME)}.
  Bonafide members appear to exist hundreds of parsecs from the core, and
  up to $\approx$10\,km\,s$^{-1}$ in tangential velocity separation.
  \label{fig:hist_physical_x_rotn}
	}
\end{figure}

%\begin{figure*}[t]
%	\begin{center}
%		\leavevmode
%		\subfloat{
%			\includegraphics[width=0.75\textwidth]{f8a.pdf}
%		}
%	
%		\subfloat{
%			\includegraphics[width=0.75\textwidth]{f8b.pdf}
%		}
%	\end{center}
%	\vspace{-0.7cm}
%	\caption{ {\bf Gaia-based components of NGC\,2516 in position and
%    velocity space, cross-matched against the rotators, in physical
%    units.}
%    {\it Top:} projections in position and 2d-tangential velocity
%    space. {\it Bottom:} radially binned versus 3d position and
%    2d velocity difference.
%    Bins have width 25\,pc in position, and 2.0\,km\,s$^{-1}$ in
%    velocity.
%		\label{fig:physical_x_rotn}
%	}
%\end{figure*}

How far away from the core, in position and velocity space, does the
halo actually extend?  We can explore this by crossmatching the
rotator sample against \nkinematic Gaia DR2 kinematic candidate
members.

Figure~\ref{fig:gaia6d_x_rotn} shows the result.  An important nuance
in interpreting this plot is that we show only stars with
$0.5<(Bp-Rp)_0<1.2$ for which our TESS pipeline successfully made
detrended light curves.  In other words---the base sample is the stars
for which we could have plausibly detected a rotation period.  Faint
stars with $Rp>16$ were beyond our light curve selection limit, and
crowded stars ({\it e.g.,} near saturated stars in the cluster center)
do not appear.  We selected the color limit above based on
Figure~\ref{fig:rot}, since the density of period detections seems to
decrease once $(Bp-Rp)_0 \gtrsim 1.25$, {\it i.e.}, for spectral types
later than $\approx$K4V.
Our expectation is that completion effects become important for
fainter stars.
At the distance of NGC\,2516, a K4V star has $T\approx 14.5$,
for which our reductions of the TESS images yield a precision of
$\approx 7$\,mmag per half-hour exposure \citep{bouma_wasp4b_2019}.
{\bf 
The AMPLITUDES of the rotation signals are like 10-100 mmag (CITE
REBULL 2020, FIGURE 10).
}

%G=18 is like Abs G 9.8, Bp-Rp not corrected of 2.7. 
% (Bp-Rp )_0 of 1.25 is (Bp-Rp) of 1.38... or Abs G of ~= 6.8.
% or G-Rp ~= 0.7
% So, G~=15... meaning Rp~=14.3.

% IN POSITION
% Bin width: 25
% Rotators: [361  82  43  30  16  18  14   8   6   2   4   0   0   0   0
% 0   0   0 0   0]
% Comparison: [495 116  65  46  28  20  22  12  11   8   5   1   0   1
% 0   2   0   0 0   0]
% Rotators < 25pc : 0.72929 +/- 0.00013
% Rotators > 25pc : 0.66172 +/- 0.00023
% ... where numerator and denom are 223/337

Qualitatively, the fraction of kinematic members with TESS-detected rotation 
periods seems like it may increase with increasing distance from the halo core---both in
position and in velocity space.
To quantify this, we computed the relelvant distances in physical coordinates.
This entailed first transforming from $(\alpha, \delta, \pi)$ to galactocentric ($X,Y,Z$) which we did 
assuming the AstroPy v4.0 coordinate standard (CITE).
We then converted the on-sky proper motions to physical units by dividing by the distance of each star.
We then computed the SEPARATIONS of all of these positions and
2d-velocities from the ``median core member'', which was selected by
taking \citetalias{cantatgaudin_gaia_2018} members with membership
probability exceeding 70\%.
Again, we only considered stars for which we could have plausibly
detected a rotation period --- the same color and observability limits
discussed above were also applied.

The results are shown in Figure~\ref{fig:physical_x_rotn}.
The halo extends to separations of about 200\,pc in physical space
from the cluster core.
This corresponds to a length of $\approx$300-350\,pc, depending on
which members of the halo are chosen as the ``tips'' on either
end.

In tangential velocity space, the fraction of stars with rotation
period detections remains high out to roughly 10\,km/s.
\citet{meingast_2021} by comparison required a physically motivated
cut in tangential velocity space of 1.5 km per sec.  This probably
improves purity, but clearly doesn't get everthing for completeness.

@@

>>>>

\title{
  Cluster Difference Imaging Photometric Survey. III.
  %The Coeval Core \& Halo of NGC\,2516
  %The Halo of NGC\,2516 Extends To At Least 300\,pc
  %An Open Cluster With a 300\,pc Halo
  An Open Cluster Spread Across 500\,pc
  %NGC\,2516  300\,pc
}

@@

% NB. the "classical tidal radius" is from e.g., Piskunov+08 or
% Kharchenko+2013. The latter givs a radius r2 of 0.71 degrees. At 380
% pc, corresponds to 4.7pc. The King tidal radius given by K13 is
% 7.70pc.
\begin{abstract}
  Recent analyses of the Gaia data have reported the existence of
  diffuse stellar populations surrounding nearby open clusters. 
  The stars in these diffuse halos could be tidal escapees from the
  cluster cores, or they could be remnants of an initially dispersed
  star formation complex.
  They could also be field stars---the false positives of overly
  sensitive clustering algorithms.
  In this study, we focus on a halo with a tip-to-tip length of
  500\,pc reported around the $\approx$150\,Myr open cluster
  NGC\,2516.
  The classical tidal radius of the cluster is $\lesssim$10\,pc.
  Combining photometry from Gaia, rotation periods from TESS, and
  lithium measurements from Gaia-ESO and GALAH, we find that the halo
  of NGC\,2516 is real and coeval with the core.
  Two in three kinematically selected halo members 25-250\,pc from the
  cluster center show rotation periods consistent with a
  gyrochronological age of 150\,Myr.
  A comparison sample of field stars shows no such trend.
  The rotation period and lithium abundances of the cluster stars are
  also anti-correlated at fixed stellar mass, as has been observed in
  similar stellar populations.
  This work quantifies the degree to which Gaia-based
  analyses can identify dispersed stellar populations against a
  background of field star contaminants, and
  expands the set of confirmed NGC\,2516 members by a factor
  of $\approx$2.
  We highlight the implications for spectroscopic survey targeting,
  open cluster dispersal, and planet searches around young stars.
\end{abstract}

\keywords{
	stellar ages (1581),
  stellar associations (1582),
  open clusters (1160),
  stellar dynamics (1596),
  stellar rotation (1629)
}

@@

Star clusters form when dense filaments in hierarchically structured
molecular clouds collide and collapse \citep{shu_star_1987}.  Over the
first 10\,Myr, feedback effects including protostellar outflows,
photoionization, radiation pressure, and supernova shocks disperse the
gas out of the cloud \citep{krumholz_star_2019}.  Since only $\sim
1\%$ of the cloud mass is converted into stars, gas dispersal enables
the majority ($\sim$90\%) of stars in the cluster to escape the
cluster's gravitational well \citep{lada_embedded_2003}.

@@

%TODO: double check this owrding.
A brief note on terminology.  Low-density stellar associations
connected to a dense population (the ``core'') have been described as
being in ``halos'', ``strings'', ``coronae'', ``snakes'',
``outskirts'', and ``tidal tails'' \citep[{\it
e.g.},][]{davenport_death_2010,kounkel_untangling_2019,roser_hyades_2019,tian_discovery_2020,meingast_2021}.
The latter term implies a particular model for the formation of the
association.  We opt for ``halo'' because it is concise and model
agnostic.  It suffers however from the suggestion of spherical
symmetry.  The halo of NGC\,2516 might be better described as ``tails'',
but the primary goal of this study is not to verify whether these
tails are tidal.

@@

\begin{figure*}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.95\textwidth]{f11.pdf}
	\end{center}
	\vspace{-0.7cm}
  \caption{ {\bf NGC\,2516 in the galactic context.} 
  Points denote halo members with $\pi/\sigma_\pi>20$.
  In our coordinate system, the Sun is at
  $\{X,Y,Z\} = \{ -8122, 0, +20.8\}\,{\rm pc}$.
  The gray line traces the past and future cluster orbit.  In
  different projections, the line spans different time intervals: the
  longest time window is visible in the $(Z,X)$ projection.
  The large blue arrows denote the median cluster velocity after
  subtracting the local standard of rest: 
  $\{v_{\rm X}, v_{\rm Y}, v_{\rm Z}\} = \{-22.2, -25.3, -4.6\}$\kms.
  Their sizes are internally proportionate.
  By comparison, $\vec{v}_{\rm LSR} = \{12.9, 245.6, 7.78\}$\kms.
  \label{fig:XYZ}
	}
\end{figure*}

@@

% FIXME FIXME FIXME
{\bf
To at least visualize the thing better...
}
Figure~\ref{fig:XYZ}.

@@

{\bf Curtis TODO: set $\mathcal{C}$ was created through a manual
curation process.}

@@

\begin{figure}[t]
	\begin{center}
		\leavevmode
		\subfloat{
			\includegraphics[width=0.49\textwidth]{f12a.pdf}
		}

		\subfloat{
			\includegraphics[width=0.49\textwidth]{f12b.pdf}
		}
	\end{center}
	\vspace{-0.7cm}
  \caption{ {\bf Projection effects for co-moving populations
  		across the sky.}
  	The map is colored by the velocity difference
  	a star co-moving with \cn\ would exhibit across
  	the equatorial sphere.
  	Actual positions of candidate \cn\ members are shown with points; the star denotes the cluster center.
	\label{fig:vproj}
	}
\end{figure}

@@

\begin{figure}[t]
	\begin{center}
			\includegraphics[width=0.49\textwidth]{f8b.pdf}
	\end{center}
	\vspace{-0.7cm}
  \caption{ {\bf Histogram of Figure~\ref{fig:physical_x_rotn}, binned
  versus 3d separation from the core and 2d velocity difference.} Bins
  have width 25\,pc in position, and 2.0\,km\,s$^{-1}$ in velocity.
  {\bf Uncertainties in the fraction are taken as the quadrature sum
  of the rel unc of the numerator and denominator (FIXME)}.  Bonafide
  members appear to exist hundreds of parsecs from the core, and up to
  $\approx$10\kms in tangential velocity separation.
  \label{fig:hist_physical_x_rotn}
	}
\end{figure}

How far away from the core, in position and velocity space, does the
halo extend?  We can explore this by crossmatching the stars with
detected rotation periods against our original target list of
\nkinematic Gaia DR2 kinematic candidate members.  For a fair
comparison, we consider only the stars with $0.5<(Bp-Rp)_0<1.2$ for
which our TESS pipeline successfully made detrended light curves.  In
other words---the base sample is the stars for which we could have
plausibly detected a rotation period.  Faint stars with $Rp>16$ were
beyond our light curve selection limit, and crowded stars ({\it e.g.,}
near saturated stars in the cluster center) are similarly omitted.  We
selected the color limit above based on Figure~\ref{fig:rot}, since
the density of period detections seems to decrease once $(Bp-Rp)_0
\gtrsim 1.25$, {\it i.e.}, for spectral types later than $\approx$K4V.
Our expectation is that completion effects become important for
fainter stars.
%
% At the distance of NGC\,2516, a K4V star has $T\approx 14.5$,
% for which our reductions of the TESS images yield a precision of
% $\approx 7$\,mmag per half-hour exposure \citep{bouma_wasp4b_2019}.
% {\bf 
% The AMPLITUDES of the rotation signals are like 10-100 mmag (CITE
% REBULL 2020, FIGURE 10).
% }
%
%G=18 is like Abs G 9.8, Bp-Rp not corrected of 2.7. 
% (Bp-Rp )_0 of 1.25 is (Bp-Rp) of 1.38... or Abs G of ~= 6.8.
% or G-Rp ~= 0.7
% So, G~=15... meaning Rp~=14.3. 
%
% IN POSITION
% Bin width: 25
% Rotators: [361  82  43  30  16  18  14   8   6   2   4   0   0   0   0
% 0   0   0 0   0]
% Comparison: [495 116  65  46  28  20  22  12  11   8   5   1   0   1
% 0   2   0   0 0   0]
% Rotators < 25pc : 0.72929 +/- 0.00013
% Rotators > 25pc : 0.66172 +/- 0.00023
%

The results are shown using physical coordinates in
Figure~\ref{fig:physical_x_rotn}.  Appendix~\ref{app:gaia6d_x_rotn}
includes a supplemental visualization in the non-transformed space of
Gaia measurements.  To compute the physical positions, we transformed
from $(\alpha, \delta, \pi)$ to galactocentric ($X,Y,Z$) assuming the
\texttt{astropy v4.0} coordinate standard \citep{astropy_2018}.  We
then converted the on-sky proper motions to physical units by dividing
by the distance of each star, taken as the inverse parallax.  The
inverse parallax approximation does not introduce significant biases
for this sample: the mean $\pi/\sigma_\pi$ and its standard deviation
are 122 and 40 respectively; the minimum is 19.5.
Qualitatively, the fraction of kinematic members with TESS-detected
rotation periods seems like it may decrease with increasing distance
from the halo core---both in position and in velocity space.

To quantify this, we computed the separation of the positions and 2D
velocities from the median core member, selected by taking the median
across \citetalias{cantatgaudin_gaia_2018} members with membership
probabilities exceeding 70\%.  The result is shown in
Figure~\ref{fig:hist_physical_x_rotn}.  The halo extends to
separations of about 200\,pc in physical space from the cluster core.
This corresponds to total a length of $\approx$300-350\,pc, depending on
which members of the halo are chosen as the ``tips'' on either end.
Broadly speaking, this agrees with the results of
\citetalias{kounkel_untangling_2019}.

In tangential velocity space, the fraction of stars with rotation
period detections remains high out to roughly 10\kms.
\citet{meingast_2021} by comparison required a physically motivated
cut in tangential velocity space of 1.5\kms.

Of course, an important aspect of the tangential velocity cut needs to
be understood in light of 
PROJECTION EFFECTS.

... So uhhhhh yeah I need to compute that map.

While this 
improves purity, it is somewhat lacking in completeness...

@@

\section{Kinematics $\otimes$ Rotation}

\begin{figure*}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.95\textwidth]{f4.pdf}
	\end{center}
	\vspace{-0.7cm}
  \caption{ {\bf Gaia-based components of NGC\,2516 in position and
  velocity space, cross-matched against the rotators.} The plotted
  stars are those with $0.5<(Bp-Rp)_0<1.2$ that meet the crowding
  restrictions described in Section~\ref{subsec:tess}: they should
  have been sufficiently bright and non-crowded to enable rotation
  period detection.  We caution that the appearance of fewer
  non-rotators being present toward the core is due the layering of
  the plot: quantitatively, stars toward the cluster center do not all
  show rotation periods in our analysis (see
  Figure~\ref{fig:hist_physical_x_rotn}).
  \label{fig:gaia6d_x_rotn}
	}
\end{figure*}

Figure~\ref{fig:gaia6d_x_rotn} is an alternative visualization of the
data shown in Figure~\ref{fig:physical_x_rotn}.  The rotation periods
in this diagram correspond to ``Set $\mathcal{A}$'' as described in
Section~\ref{subsec:tess}.  Many stars in the outer region of the
cluster show rotation periods consistent with a young age.  One issue
with the visualization as displayed however is that it hides the
amount of non-rotators toward the cluster center: we did not detect
rotation periods for roughly one in four stars at the cluster
center (see Figures~\ref{fig:physical_x_rotn}
and~\ref{fig:hist_physical_x_rotn}).
